---
title: "Bayesian mixed models with brms | ROPE"
author: Jens Roeser 
date: 'Last updated: `r Sys.Date()`'
output: 
  ioslides_presentation:
#  revealjs::revealjs_presentation:
    theme: simplex
    highlight: zenburn
    incremental: true
    transition: slower
    widescreen: true
    smaller: true

bibliography: ["references.bib"]
csl: apa.csl
link-citations: no

---

```{r setup, include=FALSE}
library(citr)
library(tidyverse)
library(magrittr)
library(lme4)
library(brms)
library(ggthemes)
library(kableExtra)
library(knitr)
library(readxl)
library(extrafont)
library(broom)
library(tidybayes)
library(janitor)
source("../scripts/functions.R")
options("kableExtra.html.bsTable" = T)
knitr::opts_chunk$set(echo = FALSE,
                      comment=NA, 
                      warning = FALSE,
                      message =FALSE)
theme_set(theme_few(base_size = 12) )
```



## ROPE


```{r echo = T}
fit_brm <- readRDS(file = "../stanout/brms_sim.rda")
beta <- posterior_samples(fit_brm, pars = "b_conditionb") %>% pull()
```

```{r echo = TRUE}
length(beta)
```

```{r echo = T}
beta[1:5]
```

```{r echo = T}
quantile(beta, probs = c(.025, .5, .975))
```


## ROPE

<div style="float: left;width: 55%">
  
- Region of practical equivalence: range of values that are practically equivalent [@kruschke2010believe; @kruschke2011bayesian].
- Point values don't represent our beliefs about what the null hypothesis is like: $H_0 = 0$
- The ROPE is the range of values that are practically equivalent to a null effect.
- Define region that is equivalent to no effect: all values in interval [-5, 5] 

</div>

<div style="float: right;width: 45%">

```{r fig.height=4.5, fig.width=4.5}
ggplot(data = NULL, aes(x = beta)) +
  geom_histogram(alpha = .55,  position = "identity", show.legend = F) +
  labs(x = bquote("Estimated effect"~hat(beta)), title = "Posterior probability distribution") 
```
</div>


## ROPE

<div style="float: left;width: 55%">

> - Region of practical equivalence: range of values that are practically equivalent [@kruschke2010believe; @kruschke2011bayesian].
> - Point values don't represent our beliefs about what the null hypothesis is like: $H_0 = 0$
> - The ROPE is the range of values that are practically equivalent to a null effect.
> - Define region that is equivalent to no effect: all values in interval [-5, 5] 

```{r echo = T}
library(bayestestR)
rope(beta, range = c(-5, 5)) 
```
</div>

<div style="float: right;width: 45%">

```{r fig.height=4.5, fig.width=4.5}
ggplot(data = NULL, aes(x = beta)) +
  geom_histogram(alpha = .55,  position = "identity", show.legend = F) +
  labs(x = bquote("Estimated effect"~hat(beta)), title = "Posterior probability distribution") 
```
</div>

  
  
## ROPE
  
<div style="float: left;width: 55%">
  
> - Region of practical equivalence: range of values that are practically equivalent [@kruschke2010believe; @kruschke2011bayesian].
> - Point values don't represent our beliefs about what the null hypothesis is like: $H_0 = 0$
> - The ROPE is the range of values that are practically equivalent to a null effect.
> - Define region that is equivalent to no effect: all values in interval [-5, 5] 

```{r echo = T}
library(bayestestR)
rope(beta, range = c(-5, 5)) 
```
</div>


<div style="float: right;width: 45%">

```{r fig.height=4.5, fig.width=4.5}
ggplot(data = NULL, aes(x = beta, fill = beta > -5 & beta < 5)) +
  geom_histogram(alpha = .55,  position = "identity", show.legend = F) +
  geom_vline(xintercept = 5, colour = "grey20", linetype = "dotted") +
  geom_vline(xintercept = -5, colour = "grey20", linetype = "dotted") +
  scale_fill_manual(values = c("grey", "darkred")) +
  labs(x = bquote("Estimated effect"~hat(beta)), title = "Posterior probability distribution") 
```
</div>


## ROPE

<div style="float: left;width: 55%">

> - Region of practical equivalence: range of values that are practically equivalent [@kruschke2010believe; @kruschke2011bayesian].
> - Point values don't represent our beliefs about what the null hypothesis is like: $H_0 = 0$
  > - The ROPE is the range of values that are practically equivalent to a null effect.
> - Define region that is equivalent to no effect: all values in interval [-Inf, 5] 

```{r echo = T}
library(bayestestR)
rope(beta, range = c(-Inf, 5))
```
</div>
  
  
  
<div style="float: left;width: 45%">
  
```{r fig.height=4.5, fig.width=4.5}
ggplot(data = NULL, aes(x = beta, fill = beta > -Inf & beta < 5)) +
  geom_histogram(alpha = .55,  position = "identity", show.legend = F) +
  geom_vline(xintercept = 5, colour = "grey20", linetype = "dotted") +
  scale_fill_manual(values = c("grey", "darkred")) +
  labs(x = bquote("Estimated effect"~hat(beta)), title = "Posterior probability distribution") 
```
</div>
  
  
  
  
## ROPE
  
>- Determine ROPE to assess the uncertainty of effect size [@makowski2019bayestestr]
>- Calculate the standardised effect size $\delta$ from posterior: $\delta = \frac{\beta}{\sigma}$ 
>- ROPE of [-0.1, 0.1] is the range of negligible effect sizes [@cohn1988statistical; @kruschke2018rejecting]. 

## ROPE


>- Determine ROPE to assess the uncertainty of effect size [@makowski2019bayestestr]
>- Calculate the standardised effect size $\delta$ from posterior: $\delta = \frac{\beta}{\sigma}$ 
>- ROPE of [-0.1, 0.1] is the range of negligible effect sizes [@cohn1988statistical; @kruschke2018rejecting]. 

```{r echo = T}
sigma <- posterior_samples(fit_brm, pars = "sigma") %>% pull()
delta <- beta / sigma
```

```{r echo = T}
mean(delta) # small effect (see Cohen's d)
```

## ROPE

>- Determine ROPE to assess the uncertainty of effect size [@makowski2019bayestestr]
>- Calculate the standardised effect size $\delta$ from posterior: $\delta = \frac{\beta}{\sigma}$ 
>- ROPE of [-0.1, 0.1] is the range of negligible effect sizes [@cohn1988statistical;@kruschke2018rejecting]. 

```{r echo = T}
rope(delta, range = c(-0.1, 0.1)) 
```

- The ROPE value indicates the extent to which the posterior cannot rule out a negligible effect. 
- A meaningful effect size should have a small proportion of posterior samples within the ROPE. 




## References

<style>
slides > slide { overflow: scroll; }
slides > slide:not(.nobackground):after {
  content: '';
}
</style>
