---
title: "Bayesian mixed models with brms | session 2"
author: Jens Roeser 
date: 'Last updated: `r Sys.Date()`'
output: 
  ioslides_presentation:
#  revealjs::revealjs_presentation:
    theme: simplex
    highlight: zenburn
    incremental: true
    transition: slower
    widescreen: true
    smaller: true

bibliography: ["references.bib"]
csl: apa.csl
link-citations: no

---

```{r setup, include=FALSE}
library(citr)
library(tidyverse)
library(magrittr)
library(lme4)
library(brms)
library(ggthemes)
library(kableExtra)
library(knitr)
library(readxl)
library(extrafont)
library(broom)
library(tidybayes)
library(janitor)
source("../scripts/functions.R")
options("kableExtra.html.bsTable" = T)
knitr::opts_chunk$set(echo = FALSE,
                      comment=NA, 
                      warning = FALSE,
                      message =FALSE)
theme_set(theme_few(base_size = 12))
```


# Outline

## Outline

- Priors
- Bayes Factors
- Distribution families: lognormal
- Reparametrisation
- ex-Gaussian and mixture models of keystroke data
- Model comparisons


# Priors

## Priors

- Prior knowledge about parameters.
- Often criticised as being subjective (which is often not true).
- Helps parameter estimation.
- Data overcome the prior (automatic Ockham's razor).
- Small samples are sensitive to priors.
- Important for Bayes Factors

## Priors

- Check defaults used before.

```{r}
fit_brm <- readRDS(file = "../stanout/brms_sim.rda")
prior_summary(fit_brm)
```


- Check what happens with more specific priors (smaller variance, negative effect)
- Reduce sample size and repeat


# Family of distributions

## Family of distributions (just some important ones)

- **Gaussian**: data are assumed to come from a normal distribution
- **Log-Normal**: zero bound data with positive skew
- **Bernoulli**: binomial data
- **Poisson**: count data
- **Zero-inflated Poisson**: count data with lots of zeros
- **Ex-Gaussian**: continuous data with positive skew
- Families for ordinal data [@burkner2019ordinal]
- Mixtures of distributions and distribution families



# Reparametrisation

## Reparametrisation

- Contrast coding
- Remove intercept
- Calculate difference (and interactions) from posterior





# Reading-while-writing data



## Recommended reading

- Bürkner's tutorials [@brms1; @brms2; @burkner2019ordinal; @burkner2019bayesian]
- Vasishth's tutorials [@sorensen2016bayesian; @nicenboim2016statistical; @vasishth2016statistical]
- Book-length intros: @gelman2020regression, @lambert2018student, @kruschke2014doing, @mcelreath2016statistical


## References

<style>
slides > slide { overflow: scroll; }
slides > slide:not(.nobackground):after {
  content: '';
}
</style>

